---
title: "Tutorial for R package `longsurvAccuracyMeasures`"
author: "Marshall Brown"
date: "3/2/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(width = 100)
```



# Tutorial 

## Load the package

Package can be downloaded directly from Github using the devtools package.

```{r eval = FALSE}
library(devtools)
###install 
devtools::install_github("mdbrown/longSurvAccuracyMeasures")
```

All package code is also available on [Github](https://github.com/mdbrown/longsurvAccuracyMeasures).

```{r}
#load libraries
library(survival)
library(tidyverse)
library(partlyconditional)
library(longsurvAccuracyMeasures)
```


### Simulated data

For this tutorial, we use data on 478 simulated observations from 100 hypothetical individuals with repeated marker measurements. 'marker_1' was simulated to be associated with the outcome `status`, while 'marker_2' is simulated to be random noise. 

```{r}
data(pc_data)

head(pc_data)
```

Note that `pc_data` is in 'long' format, with one row per measurement time. Each individual has a unique numeric subject id (`sub.id`) where event time (`time`) and event status (`status`) are repeated across marker measurement times (`meas.time`) given in months. 


### Fit a partly conditional Cox model  using the `partlyconditional` R package 

Below we use the function `PC.Cox` to fit a PC Cox model. To specify the model, we include information on patient id (`id`), survival time (`stime`), censoring status (`status`), measurement time (`measurement.time`), and markers.  Below we fit a model using raw  `meas.time` and two markers. Raw marker values are used in the model as predictors since `use.BLUP` is set to `FALSE` for both markers.  

For more information on fitting partly conditional Cox or GLM models, please see the tutorial for the R package [partlyconditional](http://rpubs.com/mdbrown/partlyconditional). 

```{r}

pc.cox.1 <-  PC.Cox(
        id = "sub.id",
        stime = "time",
        status = "status",
        measurement.time = "meas.time",  ##survival and measurement times must be on the same scale!!!
        markers = c("marker_1", "marker_2"),
        data = pc_data,
        use.BLUP = c(FALSE, FALSE), 
        knots.measurement.time = 3)

pc.cox.1


```

## Model Evaluation 

```{r}

result <- PC.evaluation(pc.cox.1,
                     newdata = pc_data,  
                     conditioning.time.window = c(18, 24), 
                     prediction.time = 12, 
                     risk.threshold = .4, 
                     pnf.threshold = .5, 
                     pcf.threshold = .18, 
                     bootstraps = 100) #should use more bootstrap replicates in practice. 

result

```


### Plot an ROC curve 

Components to plot the ROC curve can be found in `result$roc`. 

```{r}
library(ggplot2)
roc.data <- result$roc

ggplot(roc.data, aes(FPF, TPF)) + 
  geom_step(color = "dodgerblue") + 
  geom_abline(slope = 1, intercept = 0, linetype = 2)

```


## Model Calibration 

```{r}
calibration.result <- PC.calibration( pc.cox.1,
                                      newdata = pc_data,  
                                      conditioning.time.window = c(18, 24), 
                                      prediction.time = 12, 
                                      n.groups = 8, 
                                      plot = TRUE)
calibration.result$observed
calibration.result$predicted
```
